sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-15 doc/paper-jss/paper-jss.Rmd
Link to vignette

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix \ref{app:model} for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

>>>>

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}
